---
title: "NYPD Complaints"
author: "Matt Vang"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Visualizing complaints filed against NYPD police officers over the years 2007-2017

Data is provided by the [New York Civil Liberties Union](https://github.com/new-york-civil-liberties-union/NYPD-Misconduct-Complaint-Database-Updated) from 2007-2017. We will start by running the bash script to download the data. Next, use the 02_load_newyork_complaints file to clean and load it into the correct format. We can now process the data further for visualization.

## Initial Setup

Setting up necessary libraries and plots theme. Load in our RData saved from the 02 file.

```{r libraries}
library(tidyverse)
library(dplyr)
library(ggplot2)

theme_set(theme_bw())
options(repr.plot.width=4, repr.plot.height=3)

### read in new york data
load('ny_complaints.RData')
```

## Reshape data to prepare for plotting

We need to extract the number of complaints per officer information from the original dataset.

```{r create new frame, warning=FALSE}
### allegations made to each officer (collapsed multiple allegations in one incident into only ONE allegation)
officers_complaints <- ny_complaints %>%
  filter(ReceivedDate >= "2007-01-01" & ReceivedDate < "2018-01-01") %>%
  group_by(OfficerID,ComplaintID,FirstName,LastName,CurrentRankLong) %>%
  summarise() %>% ungroup()

### total number of distinct allegations made to each officer
distinct_complaints <- officers_complaints %>% group_by(OfficerID) %>%
  summarise(num_complaints = n()) %>% ungroup() %>%
  mutate(decile = ntile(num_complaints,10))
```

## Plotting

Replicating the article's graph. The top 10 percent of police officers are responsible for over 30 percent of all complaints.

```{r plot data, warning=FALSE}
distinct_complaints %>%
  group_by(decile) %>%
  summarise(total_complaints = sum(num_complaints)) %>%
  mutate(percent_of_total = total_complaints/sum(total_complaints)) %>%
  ggplot(aes(x=decile,y=percent_of_total)) + geom_histogram(stat = 'identity') +
  ylab("Proportion of complaints") + xlab("Decile")
```

